Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-24 paper.qmd
repo cleanup

However, both [renv]{.pkg} and [uv]{.pkg} have limitations. While [uv]{.pkg}
manages Python interpreter versions directly, [renv]{.pkg} does not restore the
version of [R]{.proglang} itself—installing the correct version must be done
separately using tools like [rig]{.pkg} [@rlib2023]. More critically, neither
handles system-level dependencies. If [sf]{.pkg} requires [GDAL]{.pkg} version
3.0 but the system has version 2.4 installed, neither tool can resolve this
conflict. Users must manually install system libraries, and the process differs
across operating systems.

@@

These tools represent significant progress in reproducibility within their
ecosystems, but they share a common limitation: none handles system-level
dependencies or polyglot environments. A project using [R]{.proglang},
[Python]{.proglang} [@python3], and system tools like [Quarto]{.pkg} or \LaTeX
requires coordinating multiple package managers ([renv]{.pkg} for R, [uv]{.pkg}
for Python, manual installation for system tools), each with its own
configuration and potential conflicts.

@@

- [rixpress]{.pkg} orchestrates polyglot analytical pipelines using [Nix]{.pkg}
  as the build engine. Each step runs in its own hermetic environment with
  automatic caching and dependency tracking, enabling true polyglot workflows
  across R, Python, and Julia. [Python]{.proglang} users can use
  [ryxpress]{.pkg}, a direct port.

@@

![Package repository size and freshness according to Repology. `nixpkgs` leads
in both total package count and update recency, making it well-suited for
reproducible research environments.](repology.png){#fig-repology}

@@

![Graphical representation of the polyglot pipeline. Purple nodes are
[Julia]{.proglang}, yellow nodes are [Python]{.pkg}, light blue nodes are
[R]{.proglang}, and dark blue nodes are Quarto derivations.](dag.png){#fig-dag}

@@

*   The RBC Model Simulation (`functions.jl`): This [Julia]{.proglang} script
  contains a single pure function that implements the state-space solution to
  the RBC model, based on @depaoli2009rbc. It takes the model's economic
  parameters as inputs and returns the simulated time-series data as a
  [DataFrame]{.pkg}.
*   The [XGBoost]{.pkg} Forecasting Model (`functions.py`): This
  [Python]{.proglang} script handles the complete machine learning workflow
  through a series of modular functions. Its responsibilities include feature
  engineering (creating lagged variables), splitting the data into training and
  testing sets, training the [XGBoost]{.pkg} model, and generating predictions.
*   The Visualisation (`functions.R`): This [R]{.proglang} script contains a
  function that uses [ggplot2]{.pkg} to create the final visualisation. It is
  designed to take the data frame of actual and predicted values produced by the
  Python script and generate a plot comparing the two series.

>>>>

However, both [renv]{.pkg} and [uv]{.pkg} have limitations. While [uv]{.pkg}
manages [Python]{.proglang} interpreter versions directly, [renv]{.pkg} does not
restore the version of [R]{.proglang} itself—installing the correct version must
be done separately using tools like [rig]{.pkg} [@rlib2023]. More critically,
neither handles system-level dependencies. If [sf]{.pkg} requires [GDAL]{.pkg}
version 3.0 but the system has version 2.4 installed, neither tool can resolve
this conflict. Users must manually install system libraries, and the process
differs across operating systems.

@@

These tools represent significant progress in reproducibility within their
ecosystems, but they share a common limitation: none handles system-level
dependencies or polyglot environments. A project using [R]{.proglang},
[Python]{.proglang} [@python3], and system tools like [Quarto]{.pkg} or \LaTeX
requires coordinating multiple package managers ([renv]{.pkg} for R, [uv]{.pkg}
for [Python]{.proglang}, manual installation for system tools), each with its own
configuration and potential conflicts.

@@

- [rixpress]{.pkg} orchestrates polyglot analytical pipelines using [Nix]{.pkg}
  as the build engine. Each step runs in its own hermetic environment with
  automatic caching and dependency tracking, enabling true polyglot workflows
  across [R]{.proglang}, [Python]{.proglang}, and [Julia]{.proglang}.
  [Python]{.proglang} users can use [ryxpress]{.pkg}, a direct port.

@@

![Package repository size and freshness according to Repology. `nixpkgs` leads
in both total package count and update recency, making it well-suited for
reproducible research environments.](images/repology.png){#fig-repology}

@@

![Graphical representation of the polyglot pipeline. Purple nodes are
[Julia]{.proglang}, yellow nodes are [Python]{.pkg}, light blue nodes are
[R]{.proglang}, and dark blue nodes are Quarto derivations.](images/dag.png){#fig-dag}

@@

*   The RBC Model Simulation (`functions.jl`): This [Julia]{.proglang} script
  contains a single pure function that implements the state-space solution to
  the RBC model, based on @depaoli2009rbc. It takes the model's economic
  parameters as inputs and returns the simulated time-series data as a
  [DataFrame]{.pkg}.
*   The [XGBoost]{.pkg} Forecasting Model (`functions.py`): This
  [Python]{.proglang} script handles the complete machine learning workflow
  through a series of modular functions. Its responsibilities include feature
  engineering (creating lagged variables), splitting the data into training and
  testing sets, training the [XGBoost]{.pkg} model, and generating predictions.
*   The Visualisation (`functions.R`): This [R]{.proglang} script contains a
  function that uses [ggplot2]{.pkg} to create the final visualisation. It is
  designed to take the data frame of actual and predicted values produced by the
  [Python]{.proglang} script and generate a plot comparing the two series.
